Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-22 paper/paper.tex
Address some comments

 \scw{Give an overview of the figure}
\scw{Define a convention that separates your use of natural numbers. Sometimes they
are universe levels and sometimes they are de Bruijn indices. Better to have separate
metavariables for each. (And it is possible to generalize universe levels to structures other than natural numbers. Ask Jonathan.}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
\scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
indices work.}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

\scw{motivate parallel reduction. Why are you telling us this now. Is this
  an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par}. Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

@@

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
\scw{Can we view this inductive relation as the graph of the partial function
that is defined recursively over types?}
\scw{What is this form extensible too? impredicative quantification? recursive
types? }
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

In the rest of the section, we prove four important facts about
our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
functionality (Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
property (Lemma~\ref{lemma:logrelNbackclos}).

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. \citet{nbeincoq} lacks $\eta$-law for functions. Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those extensions as part of our future work.

>>>>

%  \scw{Give an overview of the figure}
% \scw{Define a convention that separates your use of natural numbers. Sometimes they
% are universe levels and sometimes they are de Bruijn indices. Better to have separate
% metavariables for each. (And it is possible to generalize universe
% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J t a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[t]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        % \mathit{UpRen} \\
        % [[( up xi ) 0]] & := & [[0]] \\
        % [[( up xi ) Suc i]] & := & [[ Suc ren xi i ]] \\ \\
        % \mathit{Renaming ([[a < xi >]])} \\
        % [[(Pi A B) < xi >]] & := & [[Pi A < xi > B < up xi >]] \\
        % [[(a b) < xi >]] & := & [[a < xi > (# b < xi > #)]] \\
        % \ldots \\ \\
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par}. Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

@@

\label{sec:logreldep}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Void, Bool, Eq, Pi, Set, Red}
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
\scw{Many introductory texts define the relation as a recursive function
over type structure, or step-indices. You use an inductive relation instead, why?}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
\scw{Is it worth observing here that this definition is not over sets of typed
terms. That it characterizes all terms that look like booleans (i.e. evaluate to
true or false) or all terms that look like proofs (i.e. evaluate to refl). The
fact that there is no connection between p and a and b in the I-Eq case is strange
looking. Need to explain.  }
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\begin{equation*}
  \begin{split}
    [[I]] &\in [[ { j | j < i  } ->  PowerSet STm ]] \\
    [[S]] &\in [[PowerSet STm]]
  \end{split}
\end{equation*}
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} is crucial for our system to
be predicative and we will explain the reason shortly.

@@

In the rest of the section, we prove four important facts about
our logical relation: irrelevance (Lemma~\ref{lemma:logrelcoherence}),
functionality (Lemma~\ref{lemma:logreldeter}), cumulativity
(Lemma~\ref{lemma:logrelcumulativity}), and the backward closure
property (Lemma~\ref{lemma:logrelbackclos}).

@@

The dependently typed systems we have discussed so far also vary
greatly in expressiveness. For example, the Calculus of
Constructions~\citep{CoC} and MLTT without universe levels are unable
to encode large elimination. \citet{decagda} and \citet{nbeincoq} lack
identity types. % \citet{nbeincoq} lacks $\eta$-law for functions.
Among
the mechanized systems we have discussed so far, only
\citet{anand2014towards} has a full universe hierarchy.
Our object language \lang{}, while small, has a decent coverage of
features commonly seen in dependently typed languages, including
dependent pattern matching, dependent function types, a full universe
hierarchy, an intensional identity type. We hope features such as
type-directed reduction (necessary for unit $\eta$-law) and
impredicative sorts can be implemented in a similarly streamlined
fashion and we will leave those extensions as part of our future work.
